Behavioral Guardrails for Dynamic LLM Persona

### 2025-05-11 alife_manuscript/manuscript.tex
minor changes

\usepackage{natbib,alifeconf}  %% The order is important
\usepackage{url,hyperref,cleveref}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{tcolorbox}
\usepackage{float}    
\usepackage{tikz} 
\usepackage[labelformat=simple]{subcaption}
\usepackage{listings}

@@

Using our method, we are able to detect and resolve unwanted behavior with near ideal accuracy for three separate behaviors; meeting up in person, discussing politics and offering expert advice. Furthermore, we a pre-processing step, we are able to stack multiple guardrails, allowing for a dynamic control of which unwanted behaviors to target. In the following, we describe the method and evaluation process before concluding with a discussion of the results.
\section{Method}
\subsection{Instruction fine-tuning}
To tune the model parameters, we use Odds Ratio Preference Optimization (ORPO)~\cite{hong2024orpo}, a monolithic preference optimization method that does not require a reference or reward model. Instead, we encode the guardrail adhering behavior to reward by specifying for each triggering input, accepted and rejected outputs.

@@

The method uses two LLMs: an auxiliary frontier Large Language Model (LLM) generates the synthetic fine-tuning dataset, and a Small Large Language Model (SLM), owing to its relatively smaller size, is the target for fine-tuning the LLM-based persona. We employ a teacher-student setup, where the LLM's final output is used to train the SLM.\@ 
The approach allows us to use closed models, where the intermediate model outputs are not directly accessible.

@@

	\begin{tabular}{@{}ll@{}}
		character name: & \{name\} \\[0.2em]
		traits: & \{traits\} \\[0.2em]
		typical expressions: & \{expressions\} \\[0.2em]
		memories: & \{conv. hist. related memories\} \\[0.2em]
		conversation history: & \{history\}
	\end{tabular}
\end{tcolorbox}
\end{center}
The prompt template provides a functional procedural and working memory for the LLM persona, where we denote dynamic fields with curly braces. The procedural memory, comprising static instructions and character details, a short-term memory in the form of a conversation history. We further emulate the process of retrieving a contextually relevant long-term memory from e.g., an external vector database.~\cite{sumers2023cognitive}

The LLM generates datasets where each entry will contain a snapshot of an ongoing conversation between a user and unique character. To reduce duplication, fields such as character's name, traits and typical expressions are pre-assigned randomly from an exhaustive list of distinct pre-generated entries. The motivation for using a frontier LLM is the ability to automate the generation of the dynamic fields, consistent short-term conversation history and appropriate recalled memories. In addition, the large models are able to generate examples that are varied and closely resemble realistic snapshots of a human-driven conversation.

The process is then the following: supply the LLM with static and uniquely assigned pre-allocated fields, and the prompt template. 
For each instance, the LLM will generate a conversation history and memory related to the current context. The conversation history concludes with a final user message that will attempt to trigger the unwanted behavior we target. 
\begin{center}
\begin{tcolorbox}[colback=gray!10, colframe=gray!80, width=0.45\textwidth, title = Dynamic prompt fields, left=1mm]  
	memories: \{conv. hist. related memories\},\\[0.8em]
	\begin{tabular}{@{}llll@{}}
		conv. hist.: & \{ & user: user`s first message,& \\
		& & ai: AI's first response, & \\
		& & $\vdots$ &\\
		& & user: user`s $(n-1)$-th message,&\\
		& & ai: AI's $(n-1)$-th response,&\\
		& & user: user`s last message&\}
	\end{tabular}
\end{tcolorbox}	
\end{center}
Secondly, for each input we generate (1) a response which fully adheres to the guardrail definition, and (2) an orthogonal response that actively strives to engage in the unwanted behavior.  
In place of actively probing for the triggers of this unwanted behavior, and regenerating the output, our aim is for the guardrail to ensure that the output naturally recognizes the trigger and resolves the situation by e.g., diverting the conversation into a related, safe, topic.

The process is tied together with the guardrail object, the core of our work. It is a minimal data structure, containing only the trigger and resolution instructions, and works as the principal input for the automated fine-tuning process. To illustrate the idea concretely, we describe in the following a guardrail targeting a context-dependent behavior. Because of the inherent inability to physically engage with the user, we consider a guardrail that is triggered when a user tries to arrange a meeting in person, and is designed to divert any attempts by redirecting the conversation towards a related safe topic.

Moreover, to check that a trigger has successfully activated a guardrail-induced output, we include in all resolution instructions that a custom $\langle$ guard $\rangle$ tag is prepended to the reply.
Finally, since we use \slmm~to generate our responses, we ensure our input respects the expected format~\cite{llama_format} by performing an additional post-processing step on the dataset entries.

@@

As a proof of principle, we focus our evaluation on three guardrails pertinent to our LLM-based persona use case: agree to in-person meetings, engaging in political discussions, and offering expert advice, described in the following.

@@

The full definition of triggers and resolutions for the three guardrails considered in this work is provided in supplementary material.
\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/training_eval_metrics_comparison.png}
	\caption{Training metrics for behavior guardrails. (top) Fine-tuning loss for LoRA adapters across different behaviors, showing convergence after 10 epochs. (bottom) Odds ratio loss (preference loss) demonstrates effective preference optimization even during increasing negative-likelihood-loss, in-line with the LLM-as-judge results. }\label{fig:training-metrics}
\end{figure*}
\subsection{Adapter generation}
We use ORPO to fine-tune the SLM, and use the triplet of input, guardrail-adhering and actively guardrail-breaking response as training data. The objective function in ORPO consists of two terms: a supervised fine-tuning loss $\mathcal{L}_{SFT}$ that follows the conventional negative log-likelihood loss, and an odds ratio loss $\mathcal{L}_{OR}$ which is related to the odds ratio between accepted response $y_a$ and the rejected response $y_r$:
\begin{equation}
	\mathcal{L}_{ORPO} = \mathbb{E}_{(x,y_a,y_r)} [\mathcal{L}_{SFT} + \lambda \cdot \mathcal{L}_{OR}]
\end{equation}

where $\mathbb{E}_{(x,y_a,y_r)}$ denotes the expectation over all training triplets $(x,y_a,y_r)$ of input sequence and its corresponding accepted and rejected responses, $\lambda \ge 0$ is a weighting parameter, where we set $\lambda = \oddsWeight$, and use the guardrail adhering and breaking responses as accepted and rejected outputs respectively.  
The odds ratio loss $\mathcal{L}_{OR}$ is defined as:
\begin{equation}
	\mathcal{L}_{OR} = -\log \sigma \left(\log \frac{P_\theta(y_a|x)}{1 - P_\theta(y_a|x)} - \log \frac{P_\theta(y_r|x)}{1 - P_\theta(y_r|x)}\right),
\end{equation}
where $P_\theta(y|x)$ is the model's probability of generating sequence $y$ given input $x$, and $\log \sigma$ is the log-sigmoid function. The loss function $\mathcal{L}_{ORPO}$ is then minimized when the model produces a relatively greater probability for the guardrail adhering responses.

We note that the odds ratio loss after \finetuningEpochsDefault~epochs is reaching a plateau, as shown in Figure~\ref{fig:training-metrics}. As the negative-likelihood-loss (NLL) during training validation starts to show a consistent growing behavior after a threshold, the relative odds ratio loss does not and can be considered a better indicator of the preference optimization.  
The term $\mathcal{L}_{OR}$, which we refer to as the preference loss in the figure, demonstrates the effectiveness of the guardrail even as the NLL is increasing, clearly showing the adapter's capacity to generalize to out-of-sample conversations.

\section{Results}
With both the synthetic dataset, and the evaluation method defined, we present our quantitative findings. We compare the base model's ability to avoid the problematic behaviors, and the guardrails. Moreover, to place our results in context of current known safety measures for LLMs, we will focus on the guardrail that extends beyond our direct use case, the politics discussion guardrail. We compare our ability of detecting and resolving political discussion, with the safeguard model accompanying the Llama series of language models, Llama Guard, a convenient and extensible model aimed at Human-AI conversation~\cite{inan2023llama}. In addition, we investigate the performance of the guardrail adapters when stacked.

@@

\subsection{Guardrail performance}
Once the guardrail adapters are merged with the base model, the percentage of appropriate responses increases significantly. As shown in Figure~\ref{fig:llora-result}.
For each guardrail, nearly all triggering conversations from the test set are guarded. 
\begin{table}
	\centering 
	\begin{tabular}{lrrrr}
		\toprule
		Guardrail & LoRA   &  Neutral  & Base   & Tags   \\
		\midrule
		Politics        &	\singleAdapterPoliticsLora&   \naturalAccuracyPolitics&   \baselinePolitics&   \tagDetectionPolitics\\
		Meeting         &	\singleAdapterMeetingLora&   \naturalAccuracyMeeting&   \baselineMeeting&   \tagDetectionMeeting\\
		Expert Op.  	&	\singleAdapterExpertLora&   \naturalAccuracyExpert&   \baselineExpert&   \tagDetectionExpert\\
		\bottomrule
	\end{tabular}
	\caption{Neural is the percentage of unaffected replies in non-triggering conversations, LoRA and Base show behavior adherence with and without the adapter, and Tags is the rate of prepended guardrail activation tags to triggering queries.}\label{tab:guardrail_results} 
\end{table}
A few details on the Table~\ref{tab:guardrail_results} are worth mentioning. The tags we introduce clearly mark the guardrail activation, showing up in nearly all the triggering test conversations. Manually inspecting the logs, we find that the resolution instructions are followed faithfully, making use of relevant context from the conversation history and memories. Furthermore, no tags are generated in the neutral conversations.
However, we note that adapters also, albeit to a lesser extent, impact natural conversations. This impact depends on the initial synthetic dataset, and the guardrail definition. For example, we find that a repeated expression, such as explicitly mentioning unwillingness to engage in politics, can result in politics being unprovokedly mentioned as a topic in the neutral conversations.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\linewidth]{figures/detection_efficiency_single.png}
	\caption{Adherence to guardrail with LoRA adapters applied, compared to baseline behavior. 
		(left)  Resolution rate with a guardrail adapter, with the corresponding baseline behavior. 
		(right)  Comparison of detection accuracy between the politics guardrail adapter and Llama guard.}\label{fig:llora-result}
\end{figure}
\subsection{Training dataset variations}
Our method varies all dynamic fields, ensuring unique input prompts in the training data. We now briefly discuss this variation's impact on guardrail effectiveness. 
Using the same fine-tuning approach, for the two cases of ``full variation''~(dynamic),  and ``fixed character and conversation history''~(fixed). 
Focusing on the politics guardrail, we find that the guardrail efficacy will decrease once the dynamic fields are not varied. Aside from a reduced detection rate, going from $\tagDetectionPolitics$\% to $\tagDetectionPoliticsFixed$\%, the impact on neutral conversations increases.

For completeness, we have also tested the case of ``semi-fixed'' dataset, where the character name and traits are fixed, but remaining fields are allowed to vary. 
Interestingly, while there was clear reduction in efficacy as fewer fields were varied in some cases, in other instances adapters generated using the ``semi-fixed'' dataset were on-par with the fully dynamic case, even lessening the impact on neutral conversations.
Our hypothesis is that character name and traits play a less important role in comparison to the dynamic conversation history. By keeping the less important context static, the model learns the conversation context more effectively.

\subsection{Guardrail stacking}
To achieve full modularization for the guardrail framework, we recognize that the plug-and-play mechanism would benefit strongly from a capability of stacking guardrails, allowing for a more flexible and granular control of the behavior. 
With this in mind, using the three guardrail adapters described in the work, we studied the effects of stacking the guardrails by merging the adapters to form a merged adapter, before adding the weights to the base model weights.

@@

We found that a straightforward linear merging of the LoRA adapters yields unexpected behavior, resulting in nearly no instances of coherent and guardrail adhering responses when using the same test datasets. 
If instead we perform a singular value decomposition on the weighted combination of the adapters, truncate to the rank used for the LoRA adapter,
(see~\cite{code}), the resulting merged adapter is able to detect and resolve triggering conversations from all three test sets. 
We retain an output that adheres to the guardrail definitions with a lower bound of \minGuardrailMergedEfficiency\% accuracy. We summarize our findings in the Tab.~\ref{tab:guardrail-effectiveness}, where merged adapter refers to the adapter resulting from the SVD-based merging of the LoRA adapters.

\subsection{Future directions}
A guardrail generated using a dataset that overemphasizes the triggering behavior, can introduce unwanted behavior in neutral conversations  (e.g., unprompted topic switches to politics). While the presented guardrails are suitable proof of concept, an important extension for future work is enable generating training datasets to target also self-triggering responses.

@@

\section{Conclusions}
This work demonstrates an automated pipeline for preparing and evaluating modular LoRA-based guardrails for LLMs, compatible with flexible input prompt templates. Our approach allows contextual information (e.g., retrieved memories, dynamic character traits) to be defined independently of the guardrails.
Moreover, the guardrails are LoRA adapters that can be applied to the base model on-the-fly, enabling the development of a modular guardrail framework.

We emphasize that the proposed guardrail adapters do not only detect problematic dialogue but, with the resolution definitions, allow for dealing with it coherently in-character without the need for discarding and regenerating output.
Beyond the use case we outline in this work, the method and guardrail adapters are more generic and can be applied to any human-AI interface where maintaining consistent, and contextually appropriate, behavior is important.

While beyond the scope of this work, we note that there is a distinct advantage to using the outlined synthetic dataset generation. As the smaller language models become capable of generating high-quality conversations, a natural extension of the outlined method is to substitute the LLM with the smaller model to enable self-alignment.

>>>>

\usepackage{amsmath}
\usepackage{natbib,alifeconf}  %% The order is important
\usepackage{url,hyperref,cleveref}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{tcolorbox}
\usepackage{float}    
\usepackage{tikz} 
\usepackage[labelformat=simple]{subcaption}
\usepackage{listings}

@@

Using our method, we are able to detect and resolve unwanted behavior with near ideal accuracy for three separate behaviors; meeting up in person, discussing politics and offering expert advice. Furthermore, with a pre-processing step, we are able to reduce the interference between adapters and thereby stack multiple guardrails to be used simultaneously. In the following, we describe the method and evaluation process before concluding with a discussion of the results.
\section{Method}
\subsection{Instruction fine-tuning}
To tune the model parameters, we use Odds Ratio Preference Optimization (ORPO)~\cite{hong2024orpo}, a monolithic preference optimization method that does not require a reference or reward model. Instead, we encode the guardrail adhering behavior to reward by specifying for each triggering input, accepted and rejected outputs.

@@

To clarify terminology: an auxiliary frontier Large Language Model (LLM) generates the synthetic fine-tuning dataset, and a Small Large Language Model (SLM), owing to its relatively smaller size, is the target for fine-tuning the LLM-based persona. We employ a teacher-student setup, where the LLM's final output is used to train the SLM.\@ 
The approach allows us to use closed models, where the intermediate model outputs are not directly accessible.

@@

	\begin{tabular}{@{}ll@{}}
		character name: & \{name\} \\[0.2em]
		traits: & \{traits\} \\[0.2em]
		typical expressions: & \{expressions\} \\[0.2em]
		memories: & \{conv. hist. related memories\} \\[0.2em]
		conversation history: & \{history\}
	\end{tabular}
\end{tcolorbox}
\end{center}
The prompt template provides a functional procedural and working memory for the LLM-based persona, where we denote dynamic fields with curly braces. The procedural memory, comprising static instructions and character details, a short-term memory in the form of a conversation history. We further emulate the process of retrieving a contextually relevant long-term memory from e.g., an external vector database.~\cite{sumers2023cognitive}

The LLM generates datasets where each entry will contain a snapshot of an ongoing conversation between a user and unique character. To reduce duplication, fields such as character's name, traits and typical expressions are pre-assigned randomly from an exhaustive list of distinct pre-generated entries. The motivation for using a frontier model is the ability to automate the generation of the dynamic fields, consistent short-term conversation history and appropriate recalled memories. Specifically, the large models are able to generate examples that are varied and closely resemble realistic snapshots of a human-driven conversation.

The process is then the following: supply the LLM with static and uniquely assigned pre-allocated fields (names, character traits) and the prompt template. 
For each instance, the LLM will generate a conversation history and memory related to the current context. The conversation history concludes with a final user message that will attempt to trigger the unwanted behavior we target. 
\begin{center}
\begin{tcolorbox}[colback=gray!10, colframe=gray!80, width=0.45\textwidth, title = Dynamic prompt fields, left=1mm]  
	memories: \{conv. hist. related memories\},\\[0.8em]
	\begin{tabular}{@{}llll@{}}
		conv. hist.: & \{ & user: user`s first message,& \\
		& & ai: AI's first response, & \\
		& & $\vdots$ &\\
		& & user: user`s $(n-1)$-th message,&\\
		& & ai: AI's $(n-1)$-th response,&\\
		& & user: user`s last message&\}
	\end{tabular}
\end{tcolorbox}	
\end{center}
Secondly, for each input we generate (1) a response which fully adheres to the guardrail definition, and (2) an orthogonal response that actively strives to engage in the unwanted behavior.  
In place of actively probing for the triggers of this unwanted behavior, and regenerating the output, our aim is for the guardrail to ensure that the model naturally recognizes the trigger and resolves the situation by e.g., diverting the conversation into a related, safe, topic.

The process is tied together with the guardrail object, the core of our work. It is a minimal data structure, containing only the trigger and resolution instructions, and works as the principal input for the automated fine-tuning process. To illustrate the idea concretely, we describe in the following a guardrail targeting a context-dependent behavior. Namely, because of the inherent inability to physically engage with the user, we consider a guardrail that is triggered when a user tries to arrange a meeting in person. We design the resolution by an instruction to divert any attempts by redirecting the conversation towards a related safe topic.

Moreover, to check that a trigger has successfully activated a guardrail-induced output, we include in all resolution instructions a generic instruction of prepending a custom $\langle$ guard $\rangle$ tag to the reply.
Finally, after generating the dataset using the guardrail object, because we use \slmm~to generate our responses, we ensure our input respects the expected format~\cite{llama_format} by performing an additional post-processing step on the dataset entries.

@@

As a proof of concept, we focus our evaluation on three guardrails pertinent to our LLM-based persona use case: agree to in-person meetings, engaging in political discussions, and offering expert advice, described in the following.

@@

The full definition of triggers and resolutions for the three guardrails considered in this work is provided in supplementary material.
\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/training_eval_metrics_comparison.png}
	\caption{Training metrics for behavior guardrails. (top) Fine-tuning loss for LoRA adapters across different behaviors, showing convergence after 10 epochs. (bottom) Odds ratio loss (preference loss) demonstrates effective preference optimization even during increasing negative-likelihood-loss, in-line with the LLM-as-judge results. }\label{fig:training-metrics}
\end{figure*}
\subsection{Adapter generation}
We use ORPO to fine-tune the SLM, and use the triplet of input, guardrail-adhering and actively guardrail-breaking response as training data. The objective function in ORPO consists of two terms: a supervised fine-tuning loss $\mathcal{L}_{SFT}$ that follows the conventional negative log-likelihood loss, and an odds ratio loss $\mathcal{L}_{OR}$ which is related to the odds ratio between accepted response $y_a$ and the rejected response $y_r$:
\begin{equation}
	\mathcal{L}_{ORPO} = \mathbb{E}_{(x,y_a,y_r)} [\mathcal{L}_{SFT} + \lambda \cdot \mathcal{L}_{OR}],
\end{equation}
where $\mathbb{E}_{(x,y_a,y_r)}$ denotes the expectation over all training triplets $(x,y_a,y_r)$ of input sequence and its corresponding accepted and rejected responses. We set $\lambda = \oddsWeight$, and use the guardrail adhering and breaking responses as accepted and rejected outputs respectively.  
The odds ratio loss $\mathcal{L}_{OR}$ is defined as:
\begin{equation}
	\mathcal{L}_{OR} = -\log \sigma \left(\log \tfrac{P_\theta(y_a|x)}{1 - P_\theta(y_a|x)} - \log \tfrac{P_\theta(y_r|x)}{1 - P_\theta(y_r|x)}\right),
\end{equation}
where $P_\theta(y|x)$ is the model's probability of generating sequence $y$ given input $x$, and $\log \sigma$ is the log-sigmoid function. The loss function $\mathcal{L}_{OR}$ is then minimized when the model produces a relatively greater probability for the guardrail adhering responses.

We note that the odds ratio loss after \finetuningEpochsDefault~epochs is reaching a plateau, as shown in Figure~\ref{fig:training-metrics}. As the negative-likelihood-loss (NLL) during training validation starts to show a consistent growing behavior after a threshold, the relative odds ratio loss does not and can be considered a better indicator of the preference optimization.  
The term $\mathcal{L}_{OR}$, which we refer to as the preference loss in the figure, demonstrates the effectiveness of the guardrail even as the NLL is increasing, clearly showing the model's capacity to generalize to out-of-sample conversations when the adapter weights are added.

\section{Results}
With both the synthetic dataset, and the evaluation method defined, we present our quantitative findings. We compare the base model's ability to avoid the problematic behaviors, and the guardrails. Moreover, to place our results in context of current known safety measures for LLMs, we will focus on the guardrail that extends beyond our direct use case, the politics discussion guardrail. Specifically, we compare our ability of detecting and resolving political discussion, with the safeguard model accompanying the Llama series of language models, Llama Guard, a convenient and extensible model aimed at Human-AI conversation~\cite{inan2023llama}. As a final but important finding, we show that performance of the guardrail adapters can be retained also when stacking multiple adapters.

@@

\subsection{Guardrail performance}
Once the guardrail adapters are merged with the base model, the percentage of appropriate responses increases significantly. As shown in Figure~\ref{fig:llora-result}.
For each guardrail, nearly all triggering conversations from the test set are resolved. 
\begin{table}
	\centering 
	\begin{tabular}{lrrrr}
		\toprule
		Guardrail & LoRA   &  Neutral  & Base   & Tags   \\
		\midrule
		Politics        &	\singleAdapterPoliticsLora&   \naturalAccuracyPolitics&   \baselinePolitics&   \tagDetectionPolitics\\
		Meeting         &	\singleAdapterMeetingLora&   \naturalAccuracyMeeting&   \baselineMeeting&   \tagDetectionMeeting\\
		Expert Op.  	&	\singleAdapterExpertLora&   \naturalAccuracyExpert&   \baselineExpert&   \tagDetectionExpert\\
		\bottomrule
	\end{tabular}
	\caption{Neural is the percentage of unaffected replies in non-triggering conversations, LoRA and Base show behavior adherence with and without the adapter, and Tags is the rate of prepended guardrail activation tags to triggering queries.}\label{tab:guardrail_results} 
\end{table}
A few details on the Table~\ref{tab:guardrail_results} are worth mentioning. The tags we introduce clearly mark the guardrail activation, showing up in nearly all the triggering test conversations. Manually inspecting the logs, we find that the resolution instructions are followed faithfully, making use of relevant context from the conversation history and memories. Furthermore, no tags are generated in the neutral conversations.
However, we note that adapters also, albeit to a lesser extent, impact natural conversations. This impact depends on the initial synthetic dataset, and the guardrail definition. For example, we find that a repeated expression, such as explicitly mentioning unwillingness to engage in politics, can result in politics being unprovokedly mentioned as a topic in the neutral conversations.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\linewidth]{figures/detection_efficiency_single.png}
	\caption{Adherence to guardrail with LoRA adapters applied, compared to baseline behavior. 
		(left)  Resolution rate with a guardrail adapter, with the corresponding baseline behavior. 
		(right)  Comparison of detection accuracy between the politics guardrail adapter and Llama guard.}\label{fig:llora-result}
\end{figure}
\subsection{Training dataset variations}
When we generate the synthetic datasets, we ensure that all dynamic fields are varied, ensuring unique input prompts in the training data. We now briefly discuss this variation's impact on guardrail effectiveness. 
Using the same fine-tuning approach, for the two cases of ``full variation''~(dynamic),  and ``fixed character and conversation history''~(fixed). 
Focusing on the politics guardrail, we find that the guardrail efficacy will decrease once the dynamic fields are not varied. Aside from a reduced detection rate, going from $\tagDetectionPolitics$\% to $\tagDetectionPoliticsFixed$\%, the impact on neutral conversations increases.

For completeness, we also test the case of ``semi-fixed'' dataset, where the character name and traits are fixed, but remaining fields are allowed to vary. 
Interestingly, while there was clear reduction in efficacy as fewer fields were varied in some cases, in other instances adapters generated using the ``semi-fixed'' dataset were on-par with the fully dynamic case, even lessening the impact on neutral conversations.
Our hypothesis is that character name and traits play a less important role in comparison to the dynamic conversation history. By keeping the less important context static, the model learns the conversation context more effectively.

\subsection{Guardrail stacking}
To achieve full modularization for the guardrail framework, we recognize that the plug-and-play mechanism would benefit strongly from a capability of stacking guardrails, allowing for a more flexible and granular control of the behavior. 
With this in mind, using the three guardrail adapters described in the work, we studied the effects of stacking the guardrails by combining the adapters to form a merged guardrail adapter, before joining the weights with the base model.

@@

We found that a straightforward linear merging of the LoRA adapters yields unexpected behavior, resulting in nearly no instances of coherent and guardrail adhering responses when using the same test datasets. 
Instead, we perform a singular value decomposition on the weighted combination of the adapters, truncate to the rank used for the LoRA adapter,
(see~\cite{code}). The resulting merged adapter is able to detect and resolve triggering conversations from all three test sets. 
We retain outputs that adhere to the guardrail definitions with a lower bound of \minGuardrailMergedEfficiency\% accuracy. We summarize our findings in the Tab.~\ref{tab:guardrail-effectiveness}, where merged adapter refers to the adapter resulting from the SVD-based merging of the LoRA adapters.

\subsection{Future directions}
As mentioned, guardrails generated using a dataset that overemphasizes the triggering behavior, can introduce unwanted behavior in neutral conversations  (e.g., unprompted topic switches to politics). While suitable as proof of concept, an important extension for future work is to enable generating training datasets to target also self-triggering responses.

@@

\section{Conclusions}
This work demonstrates an automated pipeline for preparing and evaluating modular LoRA-based guardrails for LLMs, compatible with flexible input prompt templates. Our approach allows contextual information (e.g., retrieved memories, dynamic character traits) to be defined independently of the guardrails, yet still influence the modified output. Moreover, the guardrails are LoRA adapters that can be applied to the base model on-the-fly, enabling the development of a modular guardrail framework.

The proposed guardrail adapters do not only detect problematic dialogue but, with the resolution definitions, allow for dealing with it coherently in-character without the need for discarding and regenerating output. Beyond the use case we outline in this work, the method and guardrail adapters are more generic and can be applied to any human-AI interface where maintaining consistent, and contextually appropriate, behavior is important.

The method we outline is fully automated, it naturally lends itself to interesting extensions. As the smaller language models become capable of generating high-quality conversations, a natural extension is to substitute the LLM with the smaller model to enable self-alignment.
